Numerical modelling of the Cordilleran ice sheet

### 2014-07-20 kappa.tex
Comment-out numerical parts and deactivate todos.

% My commands
\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
\newcommand{\mref}[0]{\textbf{[ref.]}}

@@

The North American Cordillera is an immense maze of mountain ranges and
intermontane basins which once hosted an ice mass of the size of Greenland: the
Cordilleran ice sheet. Although less extensive than the former Laurentide and
Fennoscandian ice sheets, the Cordilleran ice sheet perhaps remains the least
understood Pleistocene ice sheet in terms of its extent, volume, and dynamics.
Because of the irregular topography of the ice sheet's subglacial landscape,
ice flowed along tortuous pathways, sometimes following the major topographical
troughs, and sometimes flowing across them.

This thesis presents simulations performed with PISM, a numerical ice sheet
model that approximates glacier flow by basal sliding and internal deformation.
These simulations show that the somewhat paradoxical geometry of the last
glacial maximum Cordilleran ice sheet can be largely explained by the present
distribution of temperature and precipitation in the region, provided an
accurate treatment of seasonality, orographic precipitation, and daily
temperature variability in the ice sheet model. Nevertheless, at the scale of
the last glacial cycle, this configuration appears ephemeral. In contrast, the
average stage of glaciation consists of isolated glaciers and ice caps covering
the major mountain ranges of the North American Cordillera.

The simulations presented in this thesis demonstrate the potential of combining
numerical ice sheet modelling to analysis of geological evidence in order to
improve our understanding of the former glaciations.

@@

The \emph{western margin} of the Cordilleran ice sheet covered part of the
Pacific continental shelf. An independent ice cap covering Vancouver Island
diverted flow from the main body of ice to drain into the Strait of Georgia
between Vancouver Island and the Olympic Peninsula, where it formed the Juan
de Fuca marine ice lobe, calving into the Pacific Ocean \mref. Further to the
North, the ice margin encroached the Queen Charlotte Islands. However, the
archipelago may not have been fully ice-covered, even during the maximum of
glaciation. Submarine troughs traversing several portions of the continental
shelf indicate the erosional action of additional ice streams, yet it is not
clear whether they have formed during the last glacial cycle or during earlier
glaciations \mref.

\todo{Find missing refs}

@@

\begin{table}
  \centering
  \caption{Constant parameters of the ice flow model.
           \todo{find a better reference than \citet{Aschwanden.etal.2012}
                 (library) for all the physical constants. Double-check
                 \citet{Paterson.Budd.1982} and add missing references.}}
  \label{tab:iceflowparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

@@

    $n$     & Glen exponent
            & 3
            & -
            & \mref \\

@@

    $R$     & Ideal gas constant
            & 8.31441
            & \unit{J\,mol^{-1}\,K^{-1}}
            & \mref \\

@@

    $\alpha$& \todo{check value}
            & 0.95
            & -
            & - \\

    $\delta$& \todo{find a good name}
            & 0.02
            & -
            & - \\

@@

\begin{table}
  \centering
  \caption{Constant parameters of the bedrock thermal and deformation models.
           \todo{Find out what the bedrock density corresponds to. Track down
                 original sources in \citet{Ritz.1997}}}
  \label{tab:bedrockparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

    $\rho_b$& Bedrock density
            & 3300
            & \unit{kg\,m^{-3}}
            & \mref \\

    $c_b$   & Bedrock specific heat capacity
            & 1000
            & \unit{J\,kg^{-1}\,K^{-1}}
            & \citet{Ritz.1997} \\

    $k_b$   & Bedrock thermal conductivity
            & 3.0
            & \unit{W\,m^{-1}\,K^{-1}}
            & \citet{Ritz.1997} \\

    $q_G$   & Geothermal heat flux
            & 70.0
            & \unit{mW\,m^{-2}}
            &  \\

@@

    \hline
%	pism overrides:bootstrapping geothermal flux value no var = 0.070 ;
  \end{tabular*}}
\end{table}

@@

\begin{table}
  \centering
  \caption{Constant parameters of the surface and atmosphere models.
           \todo{Find a reference for the snow and rain temperature thresholds,
                 maybe \citet{Hock.2003}.}}
  \label{tab:surfaceparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

    $T_0$   & Temperature of snow precipitation
            & 0.0
            & \degC
            & \mref \\

    $T_1$   & Temperature of rain precipitation
            & 2.0
            & \degC
            & \mref \\

@@

% ----------------------------------------------------------------------
\subsection{Numerical implementation}
% ----------------------------------------------------------------------

\todo{Add here a short section about discretization of the SIA and SSA, and
      their numerical implementation. The main subtlety, I think, lies in the
      fact that some computation need to use a ``staggered'' grid in order to
      ensure that we add SIA and SSA values corresponding to the same point in
      space. I need to check that. This section is also the place where to
      introduce geographic transformation parameters that I used throughout
      the thesis -- except in \cref{fig:map-cordillera}!}

\begin{table}
  \centering
  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
           PROJ.4 parameter names, as used in all model runs, ice volume and
           area calculations, and most figures.}
  \label{tab:projparams}
  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
    \hline
    Parameter               & Name              & Value \\
    \hline
    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
    \hline
  \end{tabular*}}
\end{table}

% ======================================================================
\section{Software implementation}
% ======================================================================

% ----------------------------------------------------------------------
\subsection{Computational workflow}
% ----------------------------------------------------------------------
\todo{Figure. computational workflow}

% ----------------------------------------------------------------------
\subsection{Contributions to PISM}
% ----------------------------------------------------------------------
\todo{Table. PISM ``issues'' identified}

% ----------------------------------------------------------------------
\subsection{Peripheral tools}
% ----------------------------------------------------------------------
A Python PDD model\\
Pre-processing scripts for GRASS GIS\\
Plotting tools for PISM\\

>>>>

% My commands
\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
\newcommand{\todo}[1]{} %\emph{[\textbf{Todo:} #1]}}
\newcommand{\mref}[0]{\textbf{[ref.]}}

@@

The North American Cordillera is an immense maze of mountain ranges and
intermontane basins which once hosted an ice mass comparable to Greenland: the
Cordilleran ice sheet. Although less extensive than the former Laurentide and
Fennoscandian ice sheets, the Cordilleran ice sheet perhaps remains the least
understood Pleistocene ice sheet in terms of its extent, volume, and dynamics.
Because of the irregular topography of the ice sheet's subglacial landscape,
ice flowed along tortuous pathways, sometimes following the major topographical
troughs, and sometimes flowing across them.

This thesis presents simulations performed with PISM, a numerical ice sheet
model that approximates glacier flow by basal sliding and internal deformation.
These simulations show that the somewhat paradoxical geometry of the last
glacial maximum Cordilleran ice sheet can be largely explained by the present
distribution of temperature and precipitation in the region combined with
paleoclimate reconstructions, provided an
accurate treatment of seasonality, orographic precipitation, and daily
temperature variability in the ice sheet model. Nevertheless, at the scale of
the last glacial cycle, this configuration appears ephemeral. In contrast, the
average stage of glaciation consists of isolated glaciers and ice caps covering
the major mountain ranges of the North American Cordillera.

The simulations presented in this thesis demonstrate the potential of combining
numerical ice sheet modelling with analysis of geological evidence in order to
improve our understanding of former glaciations.

@@

The \emph{western margin} of the Cordilleran ice sheet covered part of the
Pacific continental shelf. An independent ice cap covering Vancouver Island
diverted flow from the main body of ice to drain into the Strait of Georgia
between Vancouver Island and the Olympic Peninsula, where it formed the Juan
de Fuca marine ice lobe, calving into the Pacific Ocean. Further to the
North, the ice margin encroached the Queen Charlotte Islands. However, the
archipelago may not have been fully ice-covered, even during the maximum of
glaciation. Submarine troughs traversing several portions of the continental
shelf indicate the erosional action of additional ice streams, yet it is not
clear whether they have formed during the last glacial cycle or during earlier
glaciations.

\todo{Find refs for marine extent of the ice sheet}

@@

\begin{table}
  \centering
  \caption{Constant parameters of the ice flow model.}
  \label{tab:iceflowparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

@@

    $n$     & Glen exponent
            & 3
            & -
            & - \\

@@

    $R$     & Ideal gas constant
            & 8.31441
            & \unit{J\,mol^{-1}\,K^{-1}}
            & - \\

@@

    $\alpha$& Effective pressure coefficient (\CCLI)
            & 0.95
            & -
            & - \\

    $\delta$& Effective pressure coefficient (\CCYC)
            & 0.02
            & -
            & - \\

@@

\begin{table}
  \centering
  \caption{Constant parameters of the bedrock thermal and deformation models.}
  \label{tab:bedrockparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

    $\rho_b$& Bedrock density
            & 3300
            & \unit{kg\,m^{-3}}
            & - \\

    $c_b$   & Bedrock specific heat capacity
            & 1000
            & \unit{J\,kg^{-1}\,K^{-1}}
            & - \\

    $k_b$   & Bedrock thermal conductivity
            & 3.0
            & \unit{W\,m^{-1}\,K^{-1}}
            & - \\

    $q_G$   & Geothermal heat flux
            & 70.0
            & \unit{mW\,m^{-2}}
            & - \\

@@

@@

\begin{table}
  \centering
  \caption{Constant parameters of the surface and atmosphere models.}
  \label{tab:surfaceparams}
  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
    \hline
    Not.    & Name & Value & Unit & Source \\
    \hline

    $T_0$   & Temperature of snow precipitation
            & 0.0
            & \degC
            & - \\

    $T_1$   & Temperature of rain precipitation
            & 2.0
            & \degC
            & - \\

@@

% ----------------------------------------------------------------------
%\subsection{Numerical implementation}
% ----------------------------------------------------------------------

%\todo{Add here a short section about discretization of the SIA and SSA, and
%      their numerical implementation. The main subtlety, I think, lies in the
%      fact that some computation need to use a ``staggered'' grid in order to
%      ensure that we add SIA and SSA values corresponding to the same point in
%      space. I need to check that. This section is also the place where to
%      introduce geographic transformation parameters that I used throughout
%      the thesis -- except in \cref{fig:map-cordillera}!}

%\begin{table}
%  \centering
%  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
%           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
%           PROJ.4 parameter names, as used in all model runs, ice volume and
%           area calculations, and most figures.}
%  \label{tab:projparams}
%  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
%    \hline
%    Parameter               & Name              & Value \\
%    \hline
%    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
%    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
%    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
%    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
%    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
%    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
%    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
%    \hline
%  \end{tabular*}}
%\end{table}

% ======================================================================
%\section{Software implementation}
% ======================================================================

% ----------------------------------------------------------------------
%\subsection{Computational workflow}
% ----------------------------------------------------------------------
%\todo{Figure. computational workflow}

% ----------------------------------------------------------------------
%\subsection{Contributions to PISM}
% ----------------------------------------------------------------------
%\todo{Table. PISM ``issues'' identified}

% ----------------------------------------------------------------------
%\subsection{Peripheral tools}
% ----------------------------------------------------------------------
%A Python PDD model\\
%Pre-processing scripts for GRASS GIS\\
%Plotting tools for PISM\\
